Sequential rank agreement methods for comparison of ranked lists

### 2015-05-06 latex/paper.Rnw
Rettelser til tekst

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$. An
ordered list is a permutation function, $R: \{1,\dots,P\}\to
\{1,\dots,P\}$, such that $R(X_p)$ is the rank of item $X_p$ in the
list. The inverse mapping $\pi=R^{-1}$ assigns to rank
$r\in\{1,\dots,P\}$ the item $\pi(r)$ found at that rank. The methods
described below work for a set of $L$ lists $R_1,\dots,R_L$,
$L\geq2$. We denote $\pi_l=R_l^{-1}$ for the corresponding inverse
mappings. Panels (a) and (b) of Table~\ref{tab:example} show a
schematic example of these mappings. Thus if $\pi_l(1)=X_{34}$ then
item $X_{34}$ is ranked first in list $l$, and similarly $R_l(X_{34})=1$.

@@

The agreement of the lists regarding the rank given to an item can be
measured by
\begin{equation}
  A(p) = f(R_1(p), \ldots, R_L(p)),
\end{equation}
for a distance function $f$. Throughout this paper we will use the
sample standard error as our function $f$ and hence use
$$A(p) = \sqrt{\frac{\sum_{i=1}^L (R_i(p) - \bar{R}(p))^2}{L-1}},
$$
but other choices could be made (see the discussion). The sample
standard error has an interpretation as the average distance of the
individual rankings of the lists from the average ranking.

We now describe what is exemplified in Panel (c) of Table
\ref{tab:example} and how it can be used to define \emph{sequential
  rank agreement}. For an integer $1\le d\le P$ we define the unique
set of items found in the $L$ top $d$ parts of the lists, i.e., the
set of items ranked less than or equal to $d$ in any of the lists:
\begin{equation}
S_d = \{\pi_l(r) ; r \leq d, l = 1, \ldots, L \}.
\end{equation}
The \emph{sequential rank agreement} is the pooled standard deviation
of the items found in the set $S_d$:
\begin{equation}
\textrm{SRA}(d)= \sqrt{\frac{\sum_{\{p \in
      S_d\}}(L-1)A(p)^2}{(L-1)|S_d|}}, \label{def:sra}
\end{equation}
and small values close to zero suggests that the lists agree on the
ordering while larger values suggests disagreement. If the ranked
lists are identical then the value of SRA will be zero for all depths
$d$.  The sequential rank agreement can be interpreted as the average
distance of the individual rankings of the lists from the average
ranking for each of the items we have seen until depth $d$.

@@

If the lists generally agree on all items then we expect the
sequential rank agreement curve to be flat, but if there are changes
in the levels of rank agreement then this suggests that there are changes
in the levels of agreement. In particular, we generally expect the
agreement among the lists to be better towards the top of the lists
and worse towards the end of the lists, and in this case the sequential rank
agreement curve will start at a low level, and then have an increase
before it levels off.

@@

\subsection{Analysis of incomplete/censored lists}
Incomplete lists are also a common occurrence that arise from, for
example, comparison of top $k$ lists, when some methods only rank some
of the items (e.g., penalized regression based on the Lasso provides a
sparse set of predictors that have non-zero coefficients. There is no
obvious ordering of the predictors that have been shrunk to zero), or
if

Top $k$, methods only provide information on some ranking (Lasso),
only interested in significant ...predictors

censored

Let $\Lambda_l, l=1, \ldots, L$ be the set of items found in list $l$
so $\Lambda_l$ is the top $k_l$ list of items from list $l$ where $k_l
= |\Lambda_l|$. Note that we observe the top $k$ items for each of the
$L$ lists if $k_1=\cdots=k_L=k$. For censored lists the rank function
becomes
\begin{equation}
\tilde R_l(p) = \left\{\begin{array}{cl} \{\pi_l^{-1}(p)\} & \text{ for } p\in \Lambda_l \\ 
\{k_l+1,\dots,P\} & \text{ for } p \not\in \Lambda_l\end{array}\right.
\end{equation}
where we only know that the rank for the unobserved items in list $l$ must be larger than the largest rank observed in that list.

In the case of censored lists it is sufficient (FIXME: requires
argument) to look at depths where we have corresponding observations
so the largest rank we should consider will be
\begin{equation}
d \leq \max(k_1, \ldots, k_L).
\end{equation}

We cannot directly compute $A(p)$ for all predictors because we only
observe a censored version of $\tilde R$ for some of the
lists. Instead we assume that the rank assigned to predictor $p$ in
list $l$ is uniformly distributed among the ranks that have \emph{not}
been assigned for list $l$.

The rankings within a single list are clearly not independent since each of the lists
essentially contains full set of ranks

\begin{equation}
\tilde A(p) = \frac{\sum_{r_1; r_1\in \tilde R_1(p)}  \cdots \sum_{r_L; r_L\in \tilde R_L(p)} A(p)}{\prod_l |\tilde R_l(p)|}
\end{equation}

FIXME: If instead of running through all elements of $\tilde R_1(p)$
one would use the average rank in $\tilde R_1(p)$ we would end up with
a too small variance.

Noget med

>>>>

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$. An
ordered list is a permutation function, $R: \{X_1,\dots,X_P\}\to
\{1,\dots,P\}$, such that $R(X_p)$ is the rank of item $X_p$ in the
list. The inverse mapping $\pi=R^{-1}$ assigns to rank
$r\in\{1,\dots,P\}$ the item $\pi(r)$ found at that rank. The methods
described below work for a set of $L$ lists $R_1,\dots,R_L$,
$L\geq2$. We denote $\pi_l=R_l^{-1}$ for the corresponding inverse
mappings. Panels (a) and (b) of Table~\ref{tab:example} show a
schematic example of these mappings. Thus if $\pi_l(1)=X_{34}$ then
item $X_{34}$ is ranked first in list $l$, and similarly $R_l(X_{34})=1$.

@@

The agreement of the lists regarding the rank given to an item $X_p$ can be
measured by
\begin{equation}
  A(X_p) = f(R_1(X_p), \ldots, R_L(X_p)),
\end{equation}
for a distance function $f$. Throughout this paper we will use the
sample standard error as our function $f$ and hence use
$$A(X_p) = \sqrt{\frac{\sum_{i=1}^L (R_i(X_p) - \bar{R}(X_p))^2}{L-1}},
$$
but other choices could be made (see the discussion). The sample
standard error has an interpretation as the average distance of the
individual rankings of the lists from the average ranking.

We now describe what is exemplified in Panel (c) of Table
\ref{tab:example} and how it can be used to define \emph{sequential
  rank agreement}. For an integer $1\le d\le P$ we define the unique
set of items found in the $L$ top $d$ parts of the lists, i.e., the
set of items ranked less than or equal to $d$ in any of the lists:
\begin{equation}
S_d = \{\pi_l(r) ; r \leq d, l = 1, \ldots, L \}.
\end{equation}
The \emph{sequential rank agreement} is the pooled standard deviation
of the items found in the set $S_d$:
\begin{equation}
\textrm{SRA}(d)= \sqrt{\frac{\sum_{\{p \in
      S_d\}}(L-1)A(X_p)^2}{(L-1)|S_d|}}, \label{def:sra}
\end{equation}
and small values close to zero suggests that the lists agree on the
ordering while larger values suggests disagreement. If the ranked
lists are identical then the value of SRA will be zero for all depths
$d$.  The sequential rank agreement can be interpreted as the average
distance of the individual rankings of the lists from the average
ranking for each of the items we have seen until depth $d$.

@@

If the lists generally agree on all items then we expect the
sequential rank agreement curve to be flat, but if there are changes
in the levels of rank agreement then this suggests that there are sets
of items that the lists agree on and other sets where they rank the
items differently. In particular, we generally expect the agreement
among the lists to be better towards the top of the lists and worse
towards the end of the lists, and in this case the sequential rank
agreement curve will start at a low level, and then have an increase
before it levels off.

@@

\subsection{Analysis of incomplete/censored lists}
Incomplete lists are also a common occurrence that arise from, for
example, comparison of top $k$ lists, or when some methods only rank a
subset of the items (e.g., penalized regression based on the Lasso provides a
sparse set of predictors that have non-zero coefficients. There is no
obvious ordering of the predictors that have all been shrunk to zero).

Sequential rank agreement can be generalized to incomplete/censored
lists in the following way.  Let $\Lambda_l, l=1, \ldots, L$ be the
set of items found in list $l$ so $\Lambda_l$ is the top $k_l$ list of
items from list $l$ where $k_l = |\Lambda_l|$. Note that if
$k_1=\cdots=k_L=k$ then we observe the top $k$ items for each of the
$L$ lists. For censored lists the rank function becomes
\begin{equation}
\tilde R_l(X_p) = \left\{\begin{array}{cl} \{\pi_l^{-1}(p)\} & \text{ for } p\in \Lambda_l \\ 
\{k_l+1,\dots,P\} & \text{ for } p \not\in \Lambda_l\end{array}\right.
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.
Unless we have additional prior knowledge about the distribution of the
unranked items then we can restrict attention of the sequential rank
agreement to 
\begin{equation}
d \leq \max(k_1, \ldots, k_L),
\end{equation}
since we have no information about the ordering of the items after a
depth $d$ where all lists are censored.

The agreement, $A(X_p)$, cannot be computed directly for all
predictors because we only observe a censored version of $\tilde R$
for some of the lists. Instead we assume that the rank assigned to
item $X_p$ in list $l$ is uniformly distributed among the ranks that
have \emph{not} been assigned for list $l$,
\begin{equation}
\tilde A(X_p) = \frac{\sum_{r_1; r_1\in \tilde R_1(X_p)}  \cdots \sum_{r_L; r_L\in \tilde R_L(X_p)} A(X_p)}{\prod_l |\tilde R_l(X_p)|}.
\end{equation}

The rankings within a single list are clearly not independent since
each of the lists essentially contains full set of ranks and by

FIXME: If instead of running through all elements of $\tilde R_1(X_p)$
one would use the average rank in $\tilde R_1(X_p)$ we would end up with
a too small variance.
